---
title: "IE 5374 - Sec 1 - Group 14 - Project 1 - Task 1"
output: pdf_document
---
```{r, Libraries}
library(plotly)
library(NbClust)
library(factoextra)
library(ggplot2)
library(ClusterR)
```



Parititon data set into pre defined (k) distinct groups or clusters where each data point belongs to only one cluster.
```{r, Data1}
#Use K-means clustering methods to generate clusters

fviz_nbclust(Data1[,2:4], kmeans, method = "silhouette") #determine optimal number of clusters and plot of the number of clusters vs. the total within sum of square

Data1_km <- Data1[2:4] #selecting first three columns
km_Data1 <- kmeans(Data1[,2:4], 6) #specify number of cluster K
attributes(km_Data1) #Kmeans attributes
km_Data1$cluster #cluster values
km_Data1$centers #centroids
Data1_km$Class <- km_Data1$cluster #kmeans cluster analysis
```

Takes each observation as a separate cluster and reiterates till it identifies two observations that are closest to each other and merge the two into a cluster.
```{r, Data1}
#Use hierarchical clustering methods to generate clusters
Data1_hr <- Data1[2:4]
Data1_hr_dist <- dist(Data1_hr,method= "euclidean") #distance

Data1_hr_clust <- hclust(Data1_hr_dist, method = "complete") #Hierarchical

plot(Data1_hr_clust, cex =0.5)
rect.hclust(Data1_hr_clust, k=6, border = 2:7) #Plot cluster dendogram
Data1_km$Hr_Data1_clusters <- cutree(Data1_hr_clust, k=6)
```

Compare the results of a cluster analysis to provided class. It is mainly used to select the right clustering algorithm for a specific data set. External validation ranges between 0-1.
```{r, Data1}
#Evaluate the performance of the clustering algorithm using external validation metrics
external_validation(Data1$Class,Data1_km$Hr_Data1_clusters,method = "rand_index")
```

```{r, Data1}
#Plot (2D or 3D) the data points for each dataset and color them according to the original class
Data1_Plot_True <- plot_ly(Data1,x=~X1,y=~X2,z=~X3, color = ~as.factor(Class))

Data1_Plot_True
```

```{r, Data1}
#Plot (2D or 3D) the data points for each dataset and color them according to the class allocated by the clustering algorithm

#kmeans
Data1_Plot_km <- plot_ly(Data1,x=~X1,y=~X2,z=~X3,type = "scatter3d", color = ~as.factor(km_Data1$cluster))
Data1_Plot_km <- Data1_Plot_km %>% add_markers()

Data1_Plot_km
```

```{r, Data1}
#Plot (2D or 3D) the data points for each dataset and color them according to the class allocated by the clustering algorithm

#Hierarchical
Hr_Data1_clusters <- cutree(Data1_hr_clust, k=6)
Data1_Plot_hr <- plot_ly(Data1_hr,x=~X1,y=~X2,z=~X3,type = "scatter3d", color = ~as.factor(Hr_Data1_clusters))

Data1_Plot_hr
```
Note: As per observation, for the given data set Hierarchial clustering gives an optimum solution over rand_index.  




```{r, Data2}
#Use K-means clustering methods to generate clusters
Data2 <- na.omit(Data2)

fviz_nbclust(Data2[,2:4], kmeans, method = "wss") 

Data2_km <- Data2[2:4] 
km_Data2 <- kmeans(Data2[,2:4], 4) 
attributes(km_Data2)
km_Data2$cluster
km_Data2$centers 
Data2_km$Class <- km_Data2$cluster 
```

```{r, Data2}
#Use hierarchical clustering methods to generate clusters
Data2 <- na.omit(Data2)
Data2_hr <- Data2[1:3]
Data2_hr_dist <- dist(Data2_hr,method= "euclidean") 

Data2_hr_clust <- hclust(Data2_hr_dist, method = "complete") 


plot(Data2_hr_clust, cex =0.5)
rect.hclust(Data2_hr_clust, k=4, border = 2:7) 
Data2_km$Hr_Data2_clusters <- cutree(Data2_hr_clust, k=4)
```

```{r, Data2}
#Evaluate the performance of the clustering algorithm using external validation metrics
external_validation(Data2$Class,Data2_km$Class,method = "jaccard_index")
```

```{r, Data2}
#Plot (2D or 3D) the data points for each dataset and color them according to the original class
Data2_Plot_True <- plot_ly(Data2,x=~X,y=~Y,z=~C, color = ~as.factor(Class))

Data2_Plot_True
```

```{r, Data2}
#Plot (2D or 3D) the data points for each dataset and color them according to the class allocated by the clustering algorithm

#kmeans
Data2_Plot_km <- plot_ly(Data2,x=~X,y=~Y,z=~C,type = "scatter3d", color = ~as.factor(km_Data2$cluster))
Data2_Plot_km <- Data2_Plot_km %>% add_markers()

Data2_Plot_km
```

```{r, Data2}
#Plot (2D or 3D) the data points for each dataset and color them according to the class allocated by the clustering algorithm

#Hierarchical
Hr_Data2_clusters <- cutree(Data2_hr_clust, k=4)
Data2_Plot_hr <- plot_ly(Data2_hr,x=~X,y=~Y,z=~C,type = "scatter3d", color = ~as.factor(Hr_Data2_clusters))

Data2_Plot_hr
```





```{r, Data3}
#Use K-means clustering methods to generate clusters
Data3 <- na.omit(Data3)

fviz_nbclust(Data3[,2:4], kmeans, method = "gap_stat") 

Data3_km <- Data3[2:4] 
km_Data3 <- kmeans(Data3[,2:4], 4) 
attributes(km_Data3)
km_Data3$cluster
km_Data3$centers 
Data3_km$Class <- km_Data3$cluster 
```

```{r, Data3}
#Use hierarchical clustering methods to generate clusters
Data3 <- na.omit(Data3)
Data3_hr <- Data3[2:4]
Data3_hr_dist <- dist(Data3_hr,method= "euclidean") 

Data3_hr_clust <- hclust(Data3_hr_dist, method = "complete") 


plot(Data3_hr_clust, cex =0.5)
rect.hclust(Data3_hr_clust, k=4, border = 2:7)
Data3_km$Hr_Data3_clusters <- cutree(Data3_hr_clust, k=4)
```

```{r, Data3}
#Evaluate the performance of the clustering algorithm using external validation metrics
external_validation(Data3$Class,Data3_km$Hr_Data3_clusters,method = "rand_index")
```

```{r, Data3}
#Plot (2D or 3D) the data points for each dataset and color them according to the original class
Data3_Plot_True <- plot_ly(Data3,x=~X1,y=~X2,z=~X3, color = ~as.factor(Class))

Data3_Plot_True
```

```{r, Data3}
#Plot (2D or 3D) the data points for each dataset and color them according to the class allocated by the clustering algorithm

#kmeans
Data3_Plot_km <- plot_ly(Data3,x=~X1,y=~X2,z=~X3,type = "scatter3d", color = ~as.factor(km_Data3$cluster))
Data3_Plot_km <- Data3_Plot_km %>% add_markers()

Data3_Plot_km
```

```{r, Data3}
#Plot (2D or 3D) the data points for each dataset and color them according to the class allocated by the clustering algorithm

#Hierarchical
Hr_Data3_clusters <- cutree(Data3_hr_clust, k=4)
Data3_Plot_hr <- plot_ly(Data3_hr,x=~X1,y=~X2,z=~X3,type = "scatter3d", color = ~as.factor(Hr_Data3_clusters))

Data3_Plot_hr
```





```{r, Data4}
#Use K-means clustering methods to generate clusters
Data4 <- na.omit(Data4)

fviz_nbclust(Data4[,2:4], kmeans, method = "silhouette") 

Data4_km <- Data4[2:4] 
km_Data4 <- kmeans(Data4[,2:4], 2) 
attributes(km_Data4)
km_Data4$cluster
km_Data4$centers 
Data4_km$Class <- km_Data4$cluster 
```

```{r, Data4}
#Use hierarchical clustering methods to generate clusters
Data4 <- na.omit(Data4)
Data4_hr <- Data4[2:4]
Data4_hr_dist <- dist(Data4_hr,method= "euclidean") 

Data4_hr_clust <- hclust(Data4_hr_dist, method = "centroid") 


plot(Data4_hr_clust, cex =0.5)
rect.hclust(Data4_hr_clust, k=2, border = 2:7) 
Data4_km$Hr_Data4_clusters <- cutree(Data4_hr_clust, k=2)
```

```{r, Data4}
#Evaluate the performance of the clustering algorithm using external validation metrics
external_validation(Data4$Class,Data4_km$Hr_Data4_clusters,method = "jaccard_index")
```

```{r, Data4}
#Plot (2D or 3D) the data points for each dataset and color them according to the original class
Data4_Plot_True <- plot_ly(Data4,x=~X1,y=~X2,z=~X3, color = ~as.factor(Class))

Data4_Plot_True
```

```{r, Data4}
#Plot (2D or 3D) the data points for each dataset and color them according to the class allocated by the clustering algorithm

#kmeans
Data4_Plot_km <- plot_ly(Data4,x=~X1,y=~X2,z=~X3,type = "scatter3d", color = ~as.factor(km_Data4$cluster))
Data4_Plot_km <- Data4_Plot_km %>% add_markers()

Data4_Plot_km
```

```{r, Data4}
#Plot (2D or 3D) the data points for each dataset and color them according to the class allocated by the clustering algorithm

#Hierarchical
Hr_Data4_clusters <- cutree(Data4_hr_clust, k=2)
Data4_Plot_hr <- plot_ly(Data4_hr,x=~X1,y=~X2,z=~X3,type = "scatter3d", color = ~as.factor(Hr_Data4_clusters))

Data4_Plot_hr
```




```{r, Data5}
#Use K-means clustering methods to generate clusters
Data5 <- na.omit(Data5)

fviz_nbclust(Data5[,2:4], kmeans, method = "wss") 

Data5_km <- Data5[2:4] 
km_Data5 <- kmeans(Data5[,2:4], 2) 
attributes(km_Data5)
km_Data5$cluster
km_Data5$centers 
Data5_km$Class <- km_Data5$cluster 
```

```{r, Data5}
#Use hierarchical clustering methods to generate clusters
Data5 <- na.omit(Data5)
Data5_hr <- Data5[2:4]
Data5_hr_dist <- dist(Data5_hr,method= "euclidean") 

Data5_hr_clust <- hclust(Data5_hr_dist, method = "complete") 


plot(Data5_hr_clust, cex =0.5)
rect.hclust(Data5_hr_clust, k=2, border = 2:5) 
Data5_km$Hr_Data5_clusters <- cutree(Data5_hr_clust, k=2)
```

```{r, Data5}
#Evaluate the performance of the clustering algorithm using external validation metrics
external_validation(Data5$Class,Data5_km$Hr_Data5_clusters,method = "rand_index")
```

```{r, Data5}
#Plot (2D or 3D) the data points for each dataset and color them according to the original class
Data5_Plot_True <- plot_ly(Data5,x=~X1,y=~X2,z=~X3, color = ~as.factor(Class))

Data5_Plot_True
```

```{r, Data5}
#Plot (2D or 3D) the data points for each dataset and color them according to the class allocated by the clustering algorithm

#kmeans
Data5_Plot_km <- plot_ly(Data5,x=~X1,y=~X2,z=~X3,type = "scatter3d", color = ~as.factor(km_Data5$cluster))
Data5_Plot_km <- Data5_Plot_km %>% add_markers()

Data5_Plot_km
```

```{r, Data5}
#Plot (2D or 3D) the data points for each dataset and color them according to the class allocated by the clustering algorithm

#Hierarchical
Hr_Data5_clusters <- cutree(Data5_hr_clust, k=2)
Data5_Plot_hr <- plot_ly(Data5_hr,x=~X1,y=~X2,z=~X3,type = "scatter3d", color = ~as.factor(Hr_Data5_clusters))

Data5_Plot_hr
```





```{r, Data6}
#Use K-means clustering methods to generate clusters
Data6 <- na.omit(Data6)

fviz_nbclust(Data6[,2:3], kmeans, method = "wss") 

Data6_km <- Data6[2:3] 
km_Data6 <- kmeans(Data6[,2:3], 3) 
attributes(km_Data6)
km_Data6$cluster
km_Data6$centers 
Data6_km$Class <- km_Data6$cluster 
```

```{r, Data6}
#Use hierarchical clustering methods to generate clusters
Data6 <- na.omit(Data6)
Data6_hr <- Data6[2:3]
Data6_hr_dist <- dist(Data6_hr,method= "euclidean") 

Data6_hr_clust <- hclust(Data6_hr_dist, method = "complete") 


plot(Data6_hr_clust, cex =0.5)
rect.hclust(Data6_hr_clust, k=3, border = 2:7) 
Data6_km$Hr_Data6_clusters <- cutree(Data6_hr_clust, k=3)
```

```{r, Data6}
#Evaluate the performance of the clustering algorithm using external validation metrics
external_validation(Data6$Class,Data6_km$Class,method = "jaccard_index")
```

```{r, Data6}
#Plot (2D or 3D) the data points for each dataset and color them according to the original class
Data6_Plot_True <- plot_ly(Data6,x=~X1,y=~X2, color = ~as.factor(Class))

Data6_Plot_True
```

```{r, Data6}
#Plot (2D or 3D) the data points for each dataset and color them according to the class allocated by the clustering algorithm

#kmeans
Data6_Plot_km <- plot_ly(Data6,x=~X1,y=~X2, color = ~as.factor(km_Data6$cluster))
Data6_Plot_km <- Data6_Plot_km %>% add_markers()

Data6_Plot_km
```

```{r, Data6}
#Plot (2D or 3D) the data points for each dataset and color them according to the class allocated by the clustering algorithm

#Hierarchical
Hr_Data6_clusters <- cutree(Data6_hr_clust, k=3)
Data6_Plot_hr <- plot_ly(Data6_hr,x=~X1,y=~X2, color = ~as.factor(Hr_Data6_clusters))

Data6_Plot_hr
```



```{r, Data7}
#Use K-means clustering methods to generate clusters
Data7 <- na.omit(Data7)

fviz_nbclust(Data7[,2:3], kmeans, method = "wss") 

Data7_km <- Data7[2:3] 
km_Data7 <- kmeans(Data7[,2:3], 6) 
attributes(km_Data7)
km_Data7$cluster
km_Data7$centers 
Data7_km$Class <- km_Data7$cluster 
```

```{r, Data7}
#Use hierarchical clustering methods to generate clusters
Data7 <- na.omit(Data7)
Data7_hr <- Data7[2:3]
Data7_hr_dist <- dist(Data7_hr,method= "euclidean") 

Data7_hr_clust <- hclust(Data7_hr_dist, method = "complete") 


plot(Data7_hr_clust, cex =0.5)
rect.hclust(Data7_hr_clust, k=6, border = 2:7) 
Data7_km$Hr_Data7_clusters <- cutree(Data7_hr_clust, k=6)
```

```{r, Data7}
#Evaluate the performance of the clustering algorithm using external validation metrics
external_validation(Data7$Class,Data7_km$Class,method = "rand_index")
```

```{r, Data7}
#Plot (2D or 3D) the data points for each dataset and color them according to the original class
Data7_Plot_True <- plot_ly(Data7,x=~X1,y=~X2, color = ~as.factor(Class))

Data7_Plot_True
```

```{r, Data7}
#Plot (2D or 3D) the data points for each dataset and color them according to the class allocated by the clustering algorithm

#kmeans
Data7_Plot_km <- plot_ly(Data7,x=~X1,y=~X2, color = ~as.factor(km_Data7$cluster))
Data7_Plot_km <- Data7_Plot_km %>% add_markers()

Data7_Plot_km
```

```{r, Data7}
#Plot (2D or 3D) the data points for each dataset and color them according to the class allocated by the clustering algorithm

#Hierarchical
Hr_Data7_clusters <- cutree(Data7_hr_clust, k=6)
Data7_Plot_hr <- plot_ly(Data7_hr,x=~X1,y=~X2, color = ~as.factor(Hr_Data7_clusters))

Data7_Plot_hr
```





```{r, Data8}
#Use K-means clustering methods to generate clusters

fviz_nbclust(Data8[,2:4], kmeans, method = "silhouette") 

Data8_km <- Data8[2:4] 
km_Data8 <- kmeans(Data8[,2:4], 6) 
attributes(km_Data8)
km_Data8$cluster
km_Data8$centers 
Data8_km$Class <- km_Data8$cluster 
```

```{r, Data8}
#Use hierarchical clustering methods to generate clusters

Data8_hr <- Data8[2:4]
Data8_hr_dist <- dist(Data8_hr,method= "euclidean") 

Data8_hr_clust <- hclust(Data8_hr_dist, method = "complete") 


plot(Data8_hr_clust, cex =0.5)
rect.hclust(Data8_hr_clust, k=6, border = 2:7)
Data8_km$Hr_Data8_clusters <- cutree(Data8_hr_clust, k=6)
```

```{r, Data8}
#Evaluate the performance of the clustering algorithm using external validation metrics
external_validation(Data8$Class,Data8_km$Hr_Data8_clusters,method = "rand_index")
```

```{r, Data8}
#Plot (2D or 3D) the data points for each dataset and color them according to the original class
Data8_Plot_True <- plot_ly(Data8,x=~X1,y=~X2,z=~X3, color = ~as.factor(Class))

Data8_Plot_True
```

```{r, Data8}
#Plot (2D or 3D) the data points for each dataset and color them according to the class allocated by the clustering algorithm

#kmeans
Data8_Plot_km <- plot_ly(Data8,x=~X1,y=~X2,z=~X3, color = ~as.factor(km_Data8$cluster))
Data8_Plot_km <- Data8_Plot_km %>% add_markers()

Data8_Plot_km
```

```{r, Data8}
#Plot (2D or 3D) the data points for each dataset and color them according to the class allocated by the clustering algorithm

#Hierarchical
Hr_Data8_clusters <- cutree(Data8_hr_clust, k=6)
Data8_Plot_hr <- plot_ly(Data8_hr,x=~X1,y=~X2,z=~X3, color = ~as.factor(Hr_Data8_clusters))

Data8_Plot_hr
```